Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-03-19 VegOzone.tex
Draft for conclusions

We hypothesize that perennial vegetation will likely prioritize their growth or fast adolescence during a short, subarctic growing season and hence their carbon sequestration. Because the $\mathrm{DO_3SE}$ model currently does not simulate $A_\mathrm{n}$, we assume a first order proportionality between $A_\mathrm{n}$ and $g_\mathrm{sto}$ \citep{GCB:Medlyn2011} and effectively tune temperature and light response functions ($f_\mathrm{T}$ and $f_\mathrm{light}$) for higher $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$. The procedure is graphically illustrated in Fig.~\ref{fig:f_temp_grassland} for grassland. We aim to increase the enclosed area between probability density function (PDF) and the respective response function. We show two climatological reference periods 1990s and 2000s to indicate that the vegetation is subject to ongoing climate change and acclimation may be changing. Resulting response functions for Norway spruce and downy birch are displayed in Appendix~Figs.~\ref{fig:f_temp_spruce}--\ref{fig:f_temp_birch}.

@@

We have modeled $\mathrm{POD_y}$ with a flux threshold $y=1\,\unit{nmol\,m^{-2}\,s^{-1}}$ per projected leaf area (PLA) for three natural/semi-natural vegetation types, downy birch, Norway spruce, and perennial grassland using both MM parameterizations and a variety of bespoke response function in $f_\mathrm{T}$ and $f_\mathrm{light}$, and localized GS (described in detail in Section~\ref{subsec:do3se_parameters}). The results are shown comprehensively in Fig.~\ref{fig:pody_rel}. For comparison, critical levels for a $4\,\unit{\%}$ (silver birch), $2\,\unit{\%}$ (Norway spruce), and $10\,\unit{\%}$ (perennial grassland, above ground) reduction in biomass \citep{ICP:MappingManual2017,ESPR:Hayes2021} are shown as horizontal dashed lines. Different years are color coded as previously. Simulations with MM GS, canopy, and leaf parameters are indicated with squared markers, bespoke GS with circles. The annotation $\pm$ refers to bespoke light response function threshold ($+$ increased, $-$ decreased), while bespoke temperature response functions are indicated by categories on the x-axis. Closed/open symbols represent SWP sensitivity switched off/on in the model.

The results depicted in Fig.~\ref{fig:pody_rel} allow for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric for ozone induced damage in subarctic vegetation. In summary, we identify the following main effects on the model predicted ozone uptake with respect to the MM default parameterization:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed onset of growing season $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher light threshold $\Rightarrow$ increased/reduced uptake,
\item bespoke light threshold $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFT and between years, but the predicted ozone uptake for the bespoke parameterization is in any case larger than for the MM default parameterizations. This indicates a considerable underestimation of ozone risk for subarctic species. Drought effects (SWP) do only matter if a bespoke GS is not taken into account. We will look at the specific species or PFTs in more detail in the following.

For downy birch (Fig.~\ref{fig:pody_rel}a) all simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Downy birch displays the largest variance of ozone uptake among all species or PFTs ($10-31\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the sensitivity to the light threshold is especially pronounced, e.g. $\Delta\mathrm{POD_1} = ^{+2.5}_{-3.6}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in case of the boreal type without bespoke GS in 2019. Without bespoke GS, 2018 stands out as the year with the higher ozone damage risk than 2019. However, taking the later onset and overall shorter GS in 2018 into account, this is reversed for the boreal acclimation type. SWP induced reduction in ozone uptake is only effective in 2018 and without bespoke GS. The overall bias deduced from bespoke GS for boreal and cold with respect to MM ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019 with maximum systematic uncertainties due to the variation of the light threshold of $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

For Norway spruce (Fig.~\ref{fig:pody_rel}b), most simulations slightly exceed the CL ($0-10\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). Norway spruce is the least sensitive to acclimation in temperature and therefor the least biased. Norway spruce displays a medium sensitivity to the light threshold, e.g. $\Delta\mathrm{POD_1} = ^{+1.7}_{-1.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in case of boreal type without bespoke GS in 2018. Taking the bespoke GS into account, we find a higher ozone uptake in 2019 for all bespoke temperature response functions. The overall bias deduced from bespoke GS for boreal and cold types with respect to MM ranges between $(0.5-1.0)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.0-3.3)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019 with maximum systematic uncertainties due to the variation of the light threshold of $^{+1.0}_{-0.8}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+1.3}_{-1.0}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

@@

The estimated biomass reduction in \unit{\%} for the bespoke temperature response function of type boreal in accordance to \citet{ICP:MappingManual2017, ESPR:Hayes2021} is shown in Table~\ref{tab:biomass_reduction}. The estimated uncertainty range corresponds to the deviation given by the $\pm 20\,\unit{\%}$ variations of the light threshold and SWP=on. For downy birch and perennial grassland, substantial reductions in total biomass are predicted for any of the two years. Norway spruce is least affected. The systematic uncertainties for downy birch are one order of magnitude larger than for Norway spruce and perennial grassland. For all species or PFTs the interannual variability is larger than the systematic uncertainties estimated. In comparison with average results for the default MM parameterization (with bespoke GS), we find a non negligible (above systematic uncertainties) low-bias for both downy birch and perennial grassland. For Norway spruce, the default MM parameterization average lies within the given systematic uncertainty range, this may reflect the more comprehensive research on this species or PFT.

\begin{table}[t]
  \caption{Estimated total biomass reduction in \unit{\%} for bespoke temperature response function of type boreal with bespoke GS and SWP=off. Uncertainty ranges reported here correspond to deviations by the $\pm 20\,\unit{\%}$ variation in light threshold and SWP=on, respectively. For comparison results for MM parameterization are shown with standard deviations over both years and SWP treatments.}
  \label{tab:biomass_reduction}
\begin{tabular}{lllll}
\tophline
Year & \multicolumn{4}{c}{PFT}\\
& Downy birch & Norway spruce & \multicolumn{2}{c}{Perennial grassland}\\
\middlehline
2018 & $15.5^{+(1.9...2.1)}_{-(0.8...1.5)}$ & $2.5^{+0.2}_{-0.2}$ & $9.7^{+0.2}_{-0.2}$ & $^*12.4^{+0.2}_{-0.2}$\\
2019 & $17.4^{+2.5}_{-1.8}$ & $3.0^{+0.2}_{-0.3}$ & $10.5^{+(0.1...0.2)}_{-(0.0...0.1)}$ & $^*13.7^{+(0.1...0.3)}_{-(0.1...0.3)}$\\
\middlehline
$\left<\mathrm{MM}\right>$ & $11.2\pm 1.1$ & $2.31\pm 0.04$ & $7.5\pm 0.9$ & $^*9.6\pm 1.4$\\
\bottomhline
\end{tabular}
\belowtable{$^*$ above ground biomass} % Table Footnotes
\end{table}

\conclusions[Discussion and conclusions]
\label{sec:conc} 
\begin{itemize}
  \item What did we learn from the 2018/2019 seasons in terms of ozone damage in the far north?
  \item What are the main uncertainties in the tools we've used?
  \item What implications could this have for future scenarios in which 2018 conditions may occur more often?
\end{itemize}
PODY is no good measure without a damage function with higher temporal resolution than annual. This has been already pointed out by \citet{AE:Musselman2006}! Higher than normal spring uptake (weather conditions) does not amount to critical uptake on an annual basis. 
Weather and climatic conditioning of plants is more important for PODy than episodes of higher than usual ozone concentrations. Therefor high $\Delta \mathrm{POD_y} / \Delta t$ is not accounted for properly.
It is important to know the plant response functions to temperature, light, humidity/precipitation as best as possible if one wants to assess risk of damage.

Regarding the shift (earlier) and prolongation (later) of the growing season in the far north no general assessment can be made. Data show that in both 2018 and 2019 ozone uptake in early season was anomalous high in all studied PFTs. The extended growing season in 2019 might have countered higher episodic ozone uptake in 2018. In both years, low mid-season uptake due to weather conditions washed out any effect of these.

{\bf TODO: might be interesting to look at Mills et al 2003 and the different spatial extents of risk estimated when using AOT40 vs PODy. That could be useful to explain why we focus on PODy (rather then AOT40).- Lisa Come back to this later on in the discussions(?) - it has not been breached at Svanvik in northern Fennoscandia.}

Further stomatal conductance limit induced through ozone damage itself - varying strength and even sign (sluggishness) can shift the uptake into both directions

---
The results depicted in Fig.~\ref{fig:pody_rel} allow for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric for ozone induced damage in subarctic vegetation. In summary, we identify the following main effects on the model predicted ozone uptake with respect to the MM default parameterization:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed onset of growing season $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher light threshold $\Rightarrow$ increased/reduced uptake,
\item bespoke light threshold $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years. Thus predicted ozone uptake is XSlarger than for MM default parameterizations. This indicates a considerable underestimation of ozone risk for subarctic species.
 The impact of the timing of the GS can be at the same order of magnitude as the acclimation of response function but points into the opposite direction. Especially, the timing of $A_\mathrm{start}$ with respect to the ozone spring peak induces large uncertainties in anticipation of an acclimation to colder climates.

Bigger picture:
\begin{itemize}
\item Ozone damage assessment using mapping manual "compromise" parameterization.
\item Land-atmosphere modeling: Problems in productivity of species especially in the arctic regions due to PFTs which are not suited for the climate. Automating optimization of climate adaption of general PFTs.
\end{itemize}

GS:
The timing and trend in the SGS may differ substantially between individual species as compared to the thermal SGS though. Species-specific methods to estimate the SGS (Karlsson et al., 2018b) resulted in later SGS for European silver birch (Betula pendula) and earlier for Norway spruce (Picea abies), compared to the thermal SGS (Figure 4). The SGS of birch and spruce has also shifted to earlier in the year over time (significantly negative trend since 1994, -0.96 days year-1 and -0.98 days year-1 respectively) in northern Sweden.
Wallin et al (2013) measured gas-exchange of 40-year-old Norway spruce trees with branch cuvettes in stands at Flakaliden in northern Sweden. Measurements of light-saturated photosynthesis were made during spring recovery from dormancy during three years 2002–2004. The recovery of gas-exchange from winter dormancy was non-linear over time and the first recovery, up to 60\% of summer maximum gas-exchange, occurred already in March, more than a month before the start of growing season calculated as when the 24-h mean temperatures exceeded 5°C for five consecutive days.

Manninen et al. (2009a) summarized the results from open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch provenances were exposed to 36 – 54 ppb O3 as mean daylight concentrations. Biomass reductions of both downy birch (Betula pubescens) and mountain birch seedlings were demonstrated after one-season exposure. Moreover, the root:shoot ratio was reduced and the leaves of the mountain birch in the high O3 treatments showed necrotic stipples already in June. Mountain birch has a very rapid phenological development during spring-early summer that makes it difficult for this species to compensate for damaged leaves. It is also noteworthy that N fertilization increased the total biomass production of mountain birch markedly only under the ambient O3 both under the current climate (open-field plots) and in the about 1°C warmer climate (non-filtered air OTCs) but not under elevated O3 (Manninen et al. 2009b).
In contrast to mountain birches, northern Scots pine of local origin only showed invisible needle-level O3 impacts, such as increased respiration, decreased level of polyamines and reduced chlorophyll/carotenoid ratio and increased amount of epicuticular waxes after three years of exposure (Manninen et al. 2009a). There is no field data comparing the O3 sensitivity of birches and Scots pine from northern Europe, but both Scots pine and Norway spruce have been ranked as more O3-tolerant species than deciduous birches and Alnus incana based on crown condition in a Lithuanian study (Girgzdiene et al. 2009). Despite the relatively high O3 tolerance of Scots pine, invisible needle injuries may not be excluded in early spring conditions with high O3 concentrations though (Augustaitis and Bytnerowicz 2008, Huttunen and Manninen 2013).

@@

\dataavailability{TEXT} %% use this section when having only data sets available

@@

\authorcontribution{SF has written the manuscript, collected and processed ozone and environmental data, and performed all statistical analyses. AVV has performed on-site observation of vegetation damage induced by ozone and contributed to theory of ozone interaction with vegetation within the introduction. LE... CO... ALEB...  FS... TB... All authors contributed to discussion and conclusion of the results.} %% this section is mandatory

@@

\begin{acknowledgements}
  Bj{\o}rg Rognerud for processing SeNorge.no data with respect for begin of growing season.
  Tore Flatlandsmo Berglen for hourly pressure data from Svanhovd.
\end{acknowledgements}

>>>>

We hypothesize that perennial vegetation will likely prioritize their growth or fast adolescence during a short, subarctic growing season and hence their carbon sequestration. Because the $\mathrm{DO_3SE}$ model currently does not simulate $A_\mathrm{n}$, we assume a first order proportionality between $A_\mathrm{n}$ and $g_\mathrm{sto}$ \citep{GCB:Medlyn2011} and effectively tune temperature and light response functions ($f_\mathrm{T}$ and $f_\mathrm{light}$) for higher $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$. The procedure is graphically illustrated in Fig.~\ref{fig:f_temp_grassland} for grassland. We aim to increase the enclosed area between probability density function (PDF) and the respective response function. We show two climatological reference periods 1990s and 2000s to indicate that the vegetation is subject to ongoing climate change and acclimation may be changing. Resulting response functions for Norway spruce (Picea abies) and downy birch are displayed in Appendix~Figs.~\ref{fig:f_temp_spruce}--\ref{fig:f_temp_birch}.

@@

We have modeled $\mathrm{POD_y}$ with a flux threshold $y=1\,\unit{nmol\,m^{-2}\,s^{-1}}$ per projected leaf area (PLA) for three natural/semi-natural vegetation types, downy birch, Norway spruce, and perennial grassland using both MM parameterizations and a variety of bespoke response function in $f_\mathrm{T}$ and $f_\mathrm{light}$, and localized GS (described in detail in Section~\ref{subsec:do3se_parameters}). The results are shown comprehensively in Fig.~\ref{fig:pody_rel}. For comparison, critical levels for a $4\,\unit{\%}$ (silver birch), $2\,\unit{\%}$ (Norway spruce), and $10\,\unit{\%}$ (perennial grassland, above ground) reduction in biomass \citep{ICP:MappingManual2017,ESPR:Hayes2021} are shown as horizontal dashed lines. Different years are color coded as previously. Simulations with MM GS, canopy, and leaf parameters are indicated with squared markers, bespoke GS with circles. The annotation $\pm$ refers to bespoke light response function threshold ($+\rightarrow\mathrm{PPFD1.2}$, $-\rightarrow\mathrm{PPFD0.8}$), while bespoke temperature response functions are indicated by categories on the x-axis. Closed/open symbols represent SWP sensitivity switched off/on in the model.

The results depicted in Fig.~\ref{fig:pody_rel} allow for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric for ozone induced damage in subarctic vegetation. In summary, we identify the following main effects on the model predicted ozone uptake with respect to the MM default parameterization:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed onset of GS $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher light threshold $\Rightarrow$ increased/reduced uptake,
\item bespoke light threshold $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought effects.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years, but the predicted ozone uptake for the bespoke parameterization is in any case larger than for the MM default parameterizations and of the same order of magnitude as the variability between the years studyed here. Although our temperatue acclimations are hypothetical, this indicates a possibly considerable underestimation of ozone risk for subarctic species. Especially, an earlier GS due to climate change, that is overlapping with the ozone spring peak, has the potential to increase the ozone damage risk for vegetation with an acclimation to cold climates. Drought effects (SWP) do only matter if a bespoke GS is not taken into account. We will look at the specific species or PFTs in more detail in the following.

For downy birch (Fig.~\ref{fig:pody_rel}a) all simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Downy birch displays the largest variance of ozone uptake among all species or PFTs ($10-31\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the sensitivity to the light threshold is especially pronounced, e.g. $\Delta\mathrm{POD_1} = ^{+2.5}_{-3.6}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in case of the boreal type without bespoke GS in 2019. Without bespoke GS, 2018 displayes a higher ozone damage risk than 2019. However, taking the later onset and overall shorter GS in 2018 into account, this is reversed for the boreal acclimation type. This can be partly explained by very high light response factors (over $f_\mathrm{light}>0.8$) in the beginning of the GS in these simulations. Whether this ia a bug or a feature of the $\mathrm{DO_3SE}$ model is not yet clear. SWP induced reduction in ozone uptake is only effective in 2018 and without bespoke GS. The overall bias deduced from bespoke GS for boreal and cold with respect to MM ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019 with maximum systematic uncertainties due to the variation of the light threshold of $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

For Norway spruce (Fig.~\ref{fig:pody_rel}b), most simulations slightly exceed the CL ($0-10\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). Norway spruce is the least sensitive to an acclimation in temperature and therefor the least biased. Norway spruce displays a medium sensitivity to the light threshold, e.g. $\Delta\mathrm{POD_1} = ^{+1.7}_{-1.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in case of boreal type without bespoke GS in 2018. Taking the bespoke GS into account, we find a higher ozone uptake in 2019 for all bespoke temperature response functions. The overall bias deduced from bespoke GS for boreal and cold types with respect to MM ranges between $(0.5-1.0)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.0-3.3)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019 with maximum systematic uncertainties due to the variation of the light threshold of $^{+1.0}_{-0.8}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+1.3}_{-1.0}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

@@

The estimated biomass reduction in \unit{\%} for the bespoke temperature response function of type boreal in accordance to \citet{ICP:MappingManual2017, ESPR:Hayes2021} is shown in Table~\ref{tab:biomass_reduction}. The estimated uncertainty range corresponds to the deviation given by the $\pm 20\,\unit{\%}$ variations of the light threshold and SWP=on. For downy birch and perennial grassland, substantial reductions in total biomass are predicted for any of the two years. Norway spruce is least affected. The systematic uncertainties for downy birch are one order of magnitude larger than for Norway spruce and perennial grassland. For all species or PFTs the interannual variability is larger than the systematic uncertainties estimated for a specific acclimation and is of the same order of magnitude as the divergance between the hypothetical acclimations. For comparison, we averaged the corresponding biomass reduction for the MM temperature and light response functions with both SWP on/off but with bespoke GS over both years, we find a non negligible, which means above systematic uncertainties, low-bias for both downy birch and perennial grassland with respect to the MM default parameteraziation. For Norway spruce, the default MM parameterization average lies within the given systematic uncertainty range, this may reflect the more comprehensive research on this species or PFT.

\begin{table}[t]
  \caption{Estimated total biomass reduction in \unit{\%} for bespoke temperature response function of type boreal with bespoke GS and SWP=off. The uncertainty ranges reported here correspond to maximum a divergance by varying the light threshold and turning SWP sensitivity on in $\mathrm{DO_3SE}$. For comparison, the corresponding biomass reduction for the MM temperature and light response functions, SWP on/off, and bespoke GS averaged over both years is shown with its standard deviation.}
  \label{tab:biomass_reduction}
\begin{tabular}{lllll}
\tophline
Year & \multicolumn{4}{c}{PFT}\\
& Downy birch & Norway spruce & \multicolumn{2}{c}{Perennial grassland}\\
\middlehline
2018 & $15.5^{+(1.9...2.1)}_{-(0.8...1.5)}$ & $2.5^{+0.2}_{-0.2}$ & $9.7^{+0.2}_{-0.2}$ & $^*12.4^{+0.2}_{-0.2}$\\
2019 & $17.4^{+2.5}_{-1.8}$ & $3.0^{+0.2}_{-0.3}$ & $10.5^{+(0.1...0.2)}_{-(0.0...0.1)}$ & $^*13.7^{+(0.1...0.3)}_{-(0.1...0.3)}$\\
\middlehline
$\left<\mathrm{MM}\right>$ & $11.2\pm 1.1$ & $2.31\pm 0.04$ & $7.5\pm 0.9$ & $^*9.6\pm 1.4$\\
\bottomhline
\end{tabular}
\belowtable{$^*$ above ground biomass} % Table Footnotes
\end{table}

\conclusions[Discussion and conclusions]
\label{sec:conc}
We have examplarily studied the effect of acclimations of biomes to a subarctic climate on phyto-toxic ozone uptake. We also assessed implications on potential ozone damage risks arising at the advant of climate change. We found that conditions for ozone formation were more favourable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occured more frequently and at higher concentration levels in 2018. This was in particular caused by the extended heatwave in spring and early summer over whole Fennoscandia and associated, extensive wildfires esecially in Central Sweden (mid--end July).

The timing and trend in the onset of growing season may differ substantially between individual species in comparison to the thermal or latitudinal estimated growing season. Species-specific methods to estimate the onset \citep{IVL:Karlsson2018} show a later onset for European silver birch and earlier for Norway spruce, compared to the thermal growing season estimate. Based on MODIS data, we estimated the growing season in 2018 to be at least one month shorter than in 2019 for Norway spruce, and half a month shorter for downy birch and perennial grassland, respectively.

We have proposed a routine for gap filling of ozone data following a Reynolds's decomposition into annomalies and climatology (mulit-annual mean). We also assessed the quality of available global (ECWMF MACC and CAM reanalysis) and regional (Copernicus) near-surface ozone concentrations products and conclude that only the Copernicus product can sufficiently reproduce observed ozone concentrations in northern Fennoscandia.

Although high and more frequent ozone peak concentrations were observed in 2018, ozone concentrations were not significantly elevated by statistical means. It therefor occurs that weather conditions and constantly high ozone concentration are more important in risk assessment using flux-based metrices like $\mathrm{POD_y}$ than episodes of peak concentrations over shorter periodes of time. Consequently, accute cellular damage caused by a very high fluxes over a short periode of time is not accounted for properly in $\mathrm{POD_y}$. As pointed out by \citet{AE:Musselman2006}, a higher temporal resolution, e.g. higher susceptibility to ozone depending on leaf age, in the associated damage functions might give a more differenciated picture.

Besides a currently missing coupling to photosynthesis or nutrient cycle, ozone inflicted damage on stomatal conductance is also not accounted for in the $\mathrm{DO_3SE}$ model. Cumulative ozone uptake has been reported to reduce stomatal conductance significantly \citep[e.g.]{BGS:Lombardozzi2012} and subsequently reduces uptake of ozone. On the other hand, an ozone induced sluggishness \citep{SR:Hoshika2015} renders stomata inapt to fully close and would lead to increased ozone uptake. The presented $\mathrm{POD_y}$ may therefor be regarded as an upper limit.
{\bf TODO: might be interesting to look at Mills et al 2003 and the different spatial extents of risk estimated when using AOT40 vs PODy. That could be useful to explain why we focus on PODy (rather then AOT40).- Lisa Come back to this later on in the discussions(?) - it has not been breached at Svanvik in northern Fennoscandia.}

We defined hypothetical temperature response functions representing different levels of species assumed acclimation to cold climates. We assume the acclimation to increase the enclosed area between probability density function and the respective response function. These hypothetical species or PFTs still need varification by experiemental data. Our results, however, allowed for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric for ozone induced damage in subarctic vegetation. In particular, we identified the following main effects on the model predicted ozone uptake with respect to the default parameterization used in risk assessment:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed onset of growing season $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher light threshold $\Rightarrow$ increased/reduced uptake,
\item bespoke light threshold $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years. Predicted ozone uptake with bespoke temperature and light response function and length and timing of growing season is larger than for default parameterizations. This indicates a considerable underestimation of ozone risk for subarctic species, especially for downy birch and perennial grassland which have been studied less intensively than Norway spruce which showes a lower susceptibility to ozone \citep{Amb:Girgzdiene2009}.

Open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch were exposed elevated ozone concentrations showed biomass reductions of both downy birch and mountain birch seedlings after one-season exposure \citep{Amb:Manninen2009}. The leaves of the mountain birch in the high ozone treatments showed necrotic stipples as early as June. As mountain birch very rapidly develops during spring and early summer, compensation for damaged leaves is difficult for this species. In contrast to mountain birches, northern Scots pine of local origin only showed invisible needle-level \chem{O_3} impacts after three years of exposure. There is no field data comparing the ozone sensitivity of birches and Scots pine from northern Europe.

The impact of the timing of the growing season can be of the same order of magnitude as the acclimation of response functions but points into the opposite direction. Especially, the trend towards an earlier onset of the growing season in northern Fennoscandia with respect to the ozone spring peak induces large uncertainties for future risk assessment and includes potential higher ozone damage risk. However, the decline of ozone concentrations in arctic spring is partly caused by the removal of ozone through active vegetation at the start ot the growing season. It is therefor not clear if the peak concentrations of ozone experienced by vegetation in the future with an earlier onset of the growing season, will be reduced compared to present day. Hence, a simulated earlier onset of growing season with present day ozone concentration might overestimate ozone uptake and hence ozone induced risk of damage.

Beyond the risk assessment of ozone induced damage, out hypothetical bespoke temperature reponse function may have important indications for landsurface modeling, where problems in productivity of species especially in the arctic regions occure due to PFTs which are not suitable for the climate. An automatition of the here proposed PDF-based acclimation using machine learning techniques could overcome these issues in the future.

@@

\dataavailability{All data is available from public databases or through institutional access. $\mathrm{DO_3SE}$ modeling results can be made available upon request {\bf TODO: Upload to an archieve together with processed, gapfilled input data?}.} %% use this section when having only data sets available

@@

\authorcontribution{SF has written the manuscript, collected and processed ozone and environmental data, and performed all statistical analyses. AVV helped in framing the research articel, has performed on-site observation of vegetation damage induced by ozone, advised with regard to plant physiological processes and contributed text to the introduction. LE has contributed to discussions, setup and evaluation of $\mathrm{DO_3SE}$ modeling results, and development of the PDF-based temperature acclimation. CO has collected species or PFT parameters from literature, performed all $\mathrm{DO_3SE}$ simulations and initial graphical checks of the results. ALEB contributed with her experience regarding subartic vegetation in Finnmark. FS contributed with a summary of the 2018 meteorological conditions. TB gave guidance in a broader research sense. All authors contributed to discussions, writing, and conclusions.} %% this section is mandatory

@@

\begin{acknowledgements}
  Bj{\o}rg Rognerud for processing SeNorge.no data with respect for begin of growing season.
  Tore Flatlandsmo Berglen for hourly pressure data from Svanhovd.
  This work was financed by the Research Council of Norway (NRC) grand No. xxxxx.
\end{acknowledgements}
